---
title: "Introduction to linear models with R"
output:
  pdf_document:
    highlight: default
    toc: no
  word_document: default
  html_document:
    theme: default
    toc: yes
---

# Graphics
Simple plots in R are easy to create.
```{r}
data(iris)
plot(iris$Sepal.Length, iris$Petal.Length)
```

as are histograms.

```{r, eval=FALSE}
hist(iris$Sepal.Length)
```


The ggplot2 package is a way of easily creating beautiful plots, and which makes it easy to add colour and regression lines.

```{r, fig.width = 7, fig.height = 2.7}
library(ggplot2)
qplot(x=Sepal.Length, y=Petal.Length, data=iris, colour=Species, shape=Species)
qplot(x=Sepal.Length, y=Petal.Length, data=iris, colour=Species, shape=Species,
      geom=c('point')) + stat_smooth(method="lm")
```
```{r, fig.width = 5, fig.height = 2.5}
qplot(x=Species, y=Sepal.Length, data=iris, geom="boxplot")
```



# More on linear model syntax


**lm** is the key function used to fit models. Its usage is illustrated in the 9  case studies.

To fit a linear model using R, it is first necessary to understand the syntax for
defining models. Let’s assume that the response variable being modeled is Y and that
A, B and C are covariates that might affect Y. The table below
provides some useful examples. Note that the mathematical symbols used to define
models do not have their normal meanings!

\begin{tabular}{|l|l|}
\hline
       Syntax &                    Model\\                              
\hline
$Y\sim A     $               &   $Y = \beta_o + \beta_1 A$\\
$Y \sim -1 + A$ &                     $Y = \beta_1 A$\\
$Y\sim A+A^2$ &   	 	          $Y = \beta_o+ \beta_1( A + A^2)$\\
$Y \sim A + I(A^2)$ &                     $Y = \beta_o+ \beta_1 A + \beta_2 A^2$\\
$Y\sim A+B$  &                     $Y = \beta_o+ \beta_1 A + \beta_2 B$\\
$Y \sim A:B$  &                     $Y = \beta_o + \beta_1 AB$\\
$Y \sim A*B$ &                      $Y = \beta_o+ \beta_1 A + \beta_2 B + \beta_3 AB$\\
$Y \sim (A + B + C)^2$ &                      $Y = \beta_o+ \beta_1 A + \beta_2 B + \beta_3 C +      \beta_4AB + \beta_5AC + \beta_6 BC$
\\
\hline
\end{tabular}

Notice the difference between the third and fourth lines of the table. The function  I() is used to seperate terms, so that we get $A$ and $A^2$ included as independent variables in the regression. 

You can also include transformations. For example
```{r, eval=FALSE}
lm(log(Y)~ A + B^2+exp(C))
```
fits the model
$$\log(Y) = \beta_0 + \beta_1 A +\beta_2 B^2  + \beta_3 \exp(C) + \epsilon$$

# Trying things for yourself

You should now try your own analysis, experimenting with fitting different linear models: 

**Exercise 1**

- Load the hills dataset from the MASS library with the command
```{r}
library(MASS)
data(hills)
?hills  ### Use ? to consult the help files for a command or object.
```
- Perform some initial exploratory data analysis by producing scatter plots of each of the pairs of variables.
- Fit a linear model to predict the winning race time from the distance and the total amount of climbing.
- What are the estimated coefficients?
- What are the error estimates (standard errors) on these estimates?
- What is the deviance of the model?
- What is the $R^2$ and adjusted $R^2$ value?
- Use visreg to visualise the fitted model (look at Case Study 4 first).
- Fit the model
$$time = a + b\times dist + c\times dist^2+\epsilon.$$ 
- What are the estimated coefficients for this model? Is the additional term required, on the basis of the adjusted $R^2$ value??










